---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

### Load the data.

In this step we load the data from our working directory into a variable called activity.  

```{r Read in Data}
activity <- read.csv("activity.csv")
```

### Process/transform the data into a format suitable for your analysis

Next we convert the date field to a date data type.

```{r  Format Data}
activity$date <- as.Date(activity$date)
```


## What is mean total number of steps taken per day?

For this part of the assignment, missing values will be ignored

### Calculate the total number of steps taken per day.

We have to be sure to specify removing NA values when summing.

```{r Calculate Total}
total_per_day <- sapply(split(activity$steps, activity$date), sum, na.rm=TRUE)
```

### Make a histogram of the total number of steps taken each day.

```{r Plot Histogram, fig.height=4}
hist(total_per_day, main="Frequency of Total Steps per Day", xlab="Total Steps")
```

### Calculate and report the mean and median of the total number of steps taken per day

Again, we have to be sure to specify removing NA values. 

```{r Calcualte Mean & Median}

mean_total_steps_per_day <- mean(total_per_day, na.rm=TRUE)
median_total_steps_per_day <- median(total_per_day, na.rm=TRUE)

```

The mean of the total steps per day was `r mean_total_steps_per_day`.

The median of the total steps per day was `r median_total_steps_per_day`.

## What is the average daily activity pattern?

### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r Time Series Plot of Average Steps Per Day, fig.height=4}
mean_per_interval <- sapply(split(activity$steps, activity$interval), mean, na.rm=TRUE)
plot(names(mean_per_interval), mean_per_interval, type="l", ylab="Number of Steps", xlab="Interval")

```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r Max of Average Daily steps}

max_interval_begin <- names(mean_per_interval[mean_per_interval==max(mean_per_interval)])

max_interval_end <- as.integer(max_interval_begin) + 5

```

On average over all days, interval [`r max_interval_begin` to 
`r max_interval_end`) had the maximum number of steps.

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r get missing rows}
missing_rows <- dim(activity[is.na(activity$steps),])[1]
```

The number of rows with missing values is `r missing_rows`.

### Devise a strategy for filling in all of the missing values in the dataset.

We will fill in missing values with the median value over all days for values having the same interval as the missing value.

### Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r Create Data Set With Missing Value Filled}

##get median for each interval over all dates
median_by_interval <- sapply(split(activity$steps, activity$interval), 
                             median, na.rm=TRUE)

##copy all activities to new dataframe 
activity_no_missing <- activity

##get a vector of the rows with missing values
missing <- is.na(activity)

##loop over the intervals and filling in the missing values
## for with the median for the matching interval
for(id in 1:length(median_by_interval))
{
        activity_no_missing[missing,]$steps[
                activity_no_missing[missing,]$interval 
                        %in% as.integer(names(median_by_interval[id]))] <- 
                                median_by_interval[id]
}

```

### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r histogram and summaries for with missing values removed, fig.height=4}

total_no_missing <- 
        sapply(split(activity_no_missing$steps, activity_no_missing$date), sum)

mean_no_missing <- mean(total_no_missing)
median_no_missing <- median(total_no_missing)


hist(total_no_missing, main="Frequency of Total Steps per Day", 
        xlab="Total Steps")

```

After replacing missing values, the mean of the total steps per day was `r mean_no_missing` and the median was `r median_no_missing`.  Replacing the missing values increased the average total per day by a small amount but had no impact on the median and only slight impact in the frequency.

## Are there differences in activity patterns between weekdays and weekends?


### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).